---
title: "Stack model"
author: "Alex"
date: "2026-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

# Get cleaned data
```{r}
march_data <- read.csv("clean_march_data.csv")
# Drop row index column if present
march_data <- march_data %>% select(-any_of(c("X")))

cbb_raw <- read.csv("cbb.csv")
teams_all <- cbb_raw$TEAM
years_all <- cbb_raw$YEAR

march_data <- march_data %>% select(-TEAM)

# Convert CONF to dummy variables
march_data <- as.data.frame(model.matrix(~ . - 1, data = march_data))
```

# Split into test/train and pred/outcome
```{r}
ratio <- 0.5
set.seed(12345)

train_rows <- sample(1:nrow(march_data), ratio * nrow(march_data))
train <- march_data[train_rows, ]
test <- march_data[-train_rows, ]

test_teams <- teams_all[-train_rows]
test_years <- years_all[-train_rows]

pred_train <- train %>% select(-POSTSEASON)
outcome_train <- train$POSTSEASON

pred_test <- test %>% select(-POSTSEASON)
outcome_test <- test$POSTSEASON
```

# Scale predictors (needed for KNN, ANN, SVM)
```{r}
library(caret)

# Remove near-zero-variance columns before scaling (prevents NaN from rare dummies)
nzv <- nearZeroVar(pred_train, saveMetrics = FALSE)
if (length(nzv) > 0) {
  pred_train <- pred_train[, -nzv]
  pred_test <- pred_test[, -nzv]
}

# Scale training data and save the parameters
train_scale <- scale(pred_train)
train_center <- attr(train_scale, "scaled:center")
train_sd <- attr(train_scale, "scaled:scale")

pred_train_scaled <- as.data.frame(train_scale)
# Scale test data using training mean/sd
pred_test_scaled <- as.data.frame(scale(pred_test, center = train_center, scale = train_sd))

outcome_train_fac <- factor(outcome_train, levels = c(0, 1), labels = c("No", "Yes"))
outcome_test_fac <- factor(outcome_test, levels = c(0, 1), labels = c("No", "Yes"))

ctrl <- trainControl(method = "cv", number = 5, savePredictions = "final", classProbs = TRUE)
```

# KNN
```{r}
knn_model <- train(
  x = pred_train_scaled,
  y = outcome_train_fac,
  method = "knn",
  trControl = ctrl,
  tuneLength = 5
)
varImp(knn_model)
```

# ANN
```{r}
library(nnet)

ann_model <- train(
  x = pred_train_scaled,
  y = outcome_train_fac,
  method = "nnet",
  trControl = ctrl,
  tuneLength = 3,
  trace = FALSE,
  maxit = 200
)
varImp(ann_model)
```

# SVM
```{r}
library(kernlab)

svm_model <- train(
  x = pred_train_scaled,
  y = outcome_train_fac,
  method = "svmRadial",
  trControl = ctrl,
  tuneLength = 3
)
varImp(svm_model)
```
# Random Forest
```{r}
library(randomForest)

rf_model <- train(
  x = pred_train_scaled,
  y = outcome_train_fac,
  method = "rf",
  trControl = ctrl,
  tuneLength = 3
)
varImp(rf_model)
```

# Build meta-features
```{r}
# Get first-level predictions on training data
knn_train_preds <- predict(knn_model, pred_train_scaled, type = "prob")$Yes
ann_train_preds <- predict(ann_model, pred_train_scaled, type = "prob")$Yes
svm_train_preds <- predict(svm_model, pred_train_scaled, type = "prob")$Yes
rf_train_preds <- predict(rf_model, pred_train_scaled, type = "prob")$Yes

meta_train <- data.frame(
  knn = knn_train_preds,
  ann = ann_train_preds,
  svm = svm_train_preds,
  rf = rf_train_preds,
  outcome = outcome_train_fac
)

# Get first-level predictions on test data
knn_test_preds <- predict(knn_model, pred_test_scaled, type = "prob")$Yes
ann_test_preds <- predict(ann_model, pred_test_scaled, type = "prob")$Yes
svm_test_preds <- predict(svm_model, pred_test_scaled, type = "prob")$Yes
rf_test_preds <- predict(rf_model, pred_test_scaled, type = "prob")$Yes

meta_test <- data.frame(
  knn = knn_test_preds,
  ann = ann_test_preds,
  svm = svm_test_preds,
  rf = rf_test_preds
)
```

# Second-level stacked logistic regression
Using logistic regression as the meta-learner instead of C5.0 — C5.0 only outputs a small number of distinct leaf-node probabilities, which causes every team to clump at the same few values and leaves the Bubble tier empty. Logistic regression outputs smooth probabilities across the full 0-1 range, which is what we want for likelihood tiers.
```{r}
stack_model <- train(
  x = meta_train %>% select(-outcome),
  y = meta_train$outcome,
  method = "glm",
  family = "binomial",
  trControl = trainControl(method = "cv", number = 5, classProbs = TRUE)
)

stack_probs <- predict(stack_model, meta_test, type = "prob")$Yes
```

# Tournament likelihood per team
```{r}
# Bucket into tiers: Unlikely (0-30%), Bubble (30-70%), Likely (70-100%)
tier <- cut(
  stack_probs,
  breaks = c(-Inf, 0.30, 0.70, Inf),
  labels = c("Unlikely (0-30%)", "Bubble (30-70%)", "Likely (70-100%)"),
  right = FALSE
)

results <- data.frame(
  TEAM = test_teams,
  YEAR = test_years,
  knn      = round(meta_test$knn, 3),
  ann      = round(meta_test$ann, 3),
  svm      = round(meta_test$svm, 3),
  rf       = round(meta_test$rf, 3),
  stacked  = round(stack_probs, 3),
  tier     = tier,
  actual   = outcome_test_fac
)

# Sort by stacked probability, highest first
results <- results[order(-results$stacked), ]

table(results$tier)

results[results$tier == "Bubble (30-70%)", ]
```

# Tier accuracy — how well do the tiers match actual outcomes?
```{r}
# Cross-tab of predicted tier vs actual outcome
tier_matrix <- table(Tier = results$tier, Actual = results$actual)
tier_matrix

# Proportion of teams in each tier that actually made the tournament
tier_accuracy <- prop.table(tier_matrix, margin = 1)
round(tier_accuracy, 3)
```